---
title: "Reproducible Research - Assignment 1"
author: "Vinod"
date: "January 13, 2016"
output: html_document
---

This R Markdown document will contain all steps and corresponding code for the assignment

##Load required libraries

```{r}
library(ggplot2)
library(Hmisc)
```

##Load and preprocessing the data

```{r}
# download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "rep-research.zip")
# if(!file.exists("./repdata")) {dir.create("./repdata")}
# unzip("rep-research.zip", exdir = "./repdata")
activity <- read.csv("./repdata/activity.csv")
activity$time <- gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", activity$interval)
```

##What is mean total number of steps taken per day?

```{r}
TotalstepsPerDay <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
```

###Histogram of the total number of steps taken each day

```{r}
qplot(TotalstepsPerDay, xlab='Total steps per day', ylab='Frequency using binwith 500', binwidth=500)
```

###Calculate and report mean and median of the total number of steps taken per day

```{r}
stepsByDayMean <- mean(TotalstepsPerDay)
stepsByDayMedian <- median(TotalstepsPerDay)
```

The mean steps taken per day is `r stepsByDayMean` and median steps taken per day is `r stepsByDayMedian`.

##What is the average daily activity pattern?

```{r}
averagesteps <- aggregate(x=list(meanSteps=activity$steps), by=list(interval=activity$interval), FUN=mean, na.rm=TRUE)
```

###Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)

```{r}
ggplot(data=averagesteps, aes(x=interval, y=meanSteps)) +
geom_line() +
xlab("5-minute interval") +
ylab("average number of steps taken")
```

###Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
Maxnumbersteps <- which.max(averagesteps$meanSteps)
gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2",averagesteps[Maxnumbersteps,'interval'])
```

##Imputing missing values

###Calculate and report the total  number of days/intervals where there are missing values

```{r}
length(which(is.na(activity$steps)))
```

###Devise a strategy for filling in all of the missing values in the dataset.
###Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
Imputeddata <- activity
Imputeddata$steps <- impute(activity$steps, fun=mean)
```

###Make a histogram of the total number of steps taken each day

```{r}
ImputeddataTotalstepsPerDay <- tapply(Imputeddata$steps, Imputeddata$date, sum, na.rm=TRUE)
qplot(ImputeddataTotalstepsPerDay, xlab='Total steps per day', ylab='Frequency using binwith 500', binwidth=500)
```

###Calculate and report the mean and median total number of steps taken per day

```{r}
ImputedstepsByDayMean <- mean(ImputeddataTotalstepsPerDay)
ImputedstepsByDayMedian <- median(ImputeddataTotalstepsPerDay)
```

The mean steps taken per day is `r ImputedstepsByDayMean` and median steps taken per day is `r ImputedstepsByDayMedian` based on the Imputed data.

##Are there differences in activity patterns between weekdays and weekends?

###Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

```{r}
Imputeddata$dateType <-  ifelse(as.POSIXlt(Imputeddata$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

###Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
averagedImputeddata <- aggregate(steps ~ interval + dateType, data=Imputeddata, mean)
ggplot(averagedImputeddata, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```
